---
title: "EDA<br>Stock Prices & Volume VN30 Index Vietnam"

author: "**Pham_Dinh_Bao_Khoi** (Étudiant)<br>khoipdb.23bi14230@usth.edu.vn"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
    highlight: espresso
    code_download: TRUE
    code_folding: show
    number_sections: TRUE
    dev: "svg"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.dim = c(8, 6), fig.align = "center", out.width = "100%")
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(c('r', 'python', 'json', 'linux'), position = c('top', 'right'),
               tooltip_message = 'Click to copy', tooltip_success = 'Copied')
```

------------------------------------------------------------------------
# libraries

```{r}
# install.packages(c("tidyverse", "lubridate", "zoo", "quantmod","PerformanceAnalytics", "corrplot", "plotly","gridExtra", "scales", "TTR", "tseries"))



library(tidyverse)      # Data manipulation and visualization
library(lubridate)      # Date handling
library(zoo)            # TradingDate series
library(quantmod)       # Financial analysis
library(PerformanceAnalytics)  # Financial performance metrics
library(corrplot)       # Correlation plots
library(plotly)         # Interactive plots
library(gridExtra)      # Multiple plots
library(scales)         # Scale functions for visualization
library(TTR)            # Technical trading rules
library(tseries)  

```

```{r}
# list.files("ASTD/stock_price_VN30/") <--   CHANGE PATH TO FILE

library(readr)

list_of_files <- list.files(path = "/home/palm/ASTD/stock_price_VN30/",
                             recursive = TRUE,
                             pattern = "//.csv$",
                             full.names = TRUE)
dta <- readr::read_csv(list_of_files)

```

```{r}
# Set working directory
# Set working directory
setwd("/home/palm/ASTD/stock_price_VN30")

# Get list of all CSV files
csv_files <- list.files(pattern = "*.csv")

# Load all files into a list
stock_list <- lapply(csv_files, function(file) {
  df <- read.csv(file, stringsAsFactors = FALSE)
  df$Company <- tools::file_path_sans_ext(file)  # Add company name
  return(df)
})

# Combine all stocks into one dataframe
all_stocks <- bind_rows(stock_list)

# View the structure
str(all_stocks)
head(all_stocks)

```
```{r}
# Example: Load one stock for detailed analysis
stock_sample <- read.csv("/home/palm/ASTD/stock_price_VN30/MSN.csv")

head(stock_sample)


```
## basic info

```{r}
# Dimensions
dim(all_stocks)
dim(stock_sample)

# Column names
colnames(all_stocks)

# Data types
str(all_stocks)

# First and last rows
head(all_stocks, 10)
tail(all_stocks, 10)

# Summary statistics
summary(all_stocks)

# Unique companies
unique(all_stocks$Company)
length(unique(all_stocks$Company))


```
## check Missing values
```{r}
# Missing values per column
colSums(is.na(all_stocks))

# Percentage of missing values
missing_pct <- colMeans(is.na(all_stocks)) * 100
print(missing_pct) # it show blank figure

# Visualize missing values
library(naniar)
vis_miss(all_stocks)

# Missing values by company
all_stocks %>%
  group_by(Company) %>%
  summarise(
    total_rows = n(),
    missing_close = sum(is.na(Close)),
    missing_volume = sum(is.na(Volume))
  )

```
## Check for Duplicates
however, we can easily see that, there are no uplicated values
this dataset so CLEAN !!

```{r}
# Check for duplicate rows
duplicates <- all_stocks %>%
  group_by(Company, TradingDate) %>%
  filter(n() > 1)

nrow(duplicates)

# Remove duplicates if any
all_stocks <- all_stocks %>%
  distinct(Company, TradingDate, .keep_all = TRUE)
```


# Data Cleaning and Preprocessing

## Cleaning missing values
```{r}
all_stocks <- all_stocks %>%
  select(where(~ !all(is.na(.))))

```

## DAte conversion

```{r}
format(Sys.time())
```

```{r}
# Convert Date column to Date type
all_stocks$TradingDate <- dmy(all_stocks$TradingDate)

# Sort by company and date
all_stocks <- all_stocks %>%
  arrange(Company, TradingDate)

# Check date range
range(all_stocks$TradingDate, na.rm = TRUE)
```




## Create Additional Features

explain how this section works, why it's necessary

```{r}
# Daily returns (percentage change)
all_stocks <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  mutate(
    Daily_Return = (Close - lag(Close)) / lag(Close) * 100,
    Log_Return = log(Close / lag(Close)),
    Price_Change = Close - lag(Close),
    Volume_Change = Volume - lag(Volume),
    High_Low_Range = High - Low,
    High_Low_Pct = (High - Low) / Low * 100
  ) %>%
  ungroup()

# Moving averages
all_stocks <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  mutate(
    MA_7 = rollmean(Close, k = 7, fill = NA, align = "right"),
    MA_30 = rollmean(Close, k = 30, fill = NA, align = "right"),
    MA_90 = rollmean(Close, k = 90, fill = NA, align = "right")
  ) %>%
  ungroup()

# Volatility (rolling standard deviation)
all_stocks <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  mutate(
    Volatility_30 = rollapply(Daily_Return, width = 30, 
                               FUN = sd, fill = NA, align = "right")
  ) %>%
  ungroup()

# Time-based features
all_stocks <- all_stocks %>%
  mutate(
    Year = year(TradingDate),
    Month = month(TradingDate, label = TRUE),
    Quarter = quarter(TradingDate),
    DayOfWeek = wday(TradingDate, label = TRUE),
    WeekOfYear = week(TradingDate)
  )

```

# Univariate Analysis


## Distribution of Closing Prices" top 5 companies with highest closing prices

Closing price in what time? Is this an avg of all time?

From this box plot, what is the insight? what information we can understand from these figures?

The Density plot is too wide/ small, how to make it more clearly?

Showing Distribution of 1 company for instance.

```{r}
# Identify top 5 companies by average closing price
top5_by_price <- all_stocks %>%
  group_by(Company) %>%
  summarise(Avg_Close = mean(Close, na.rm = TRUE)) %>%
  arrange(desc(Avg_Close)) %>%
  head(5) %>%
  pull(Company)

# Display the top 5 companies
cat("Top 5 Companies by Average Closing Price:\n")
top5_summary <- all_stocks %>%
  filter(Company %in% top5_by_price) %>%
  group_by(Company) %>%
  summarise(
    Avg_Close = mean(Close, na.rm = TRUE),
    Max_Close = max(Close, na.rm = TRUE),
    Min_Close = min(Close, na.rm = TRUE)
  ) %>%
  arrange(desc(Avg_Close))
print(top5_summary)

# Filter data for top 5 companies only
top5_data <- all_stocks %>%
  filter(Company %in% top5_by_price)

# Histogram - Top 5 companies only
ggplot(top5_data, aes(x = Close)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  facet_wrap(~Company, scales = "free") +
  labs(title = "Distribution of Closing Prices - Top 5 Companies by Price",
       subtitle = "Top 5 companies with highest average closing prices",
       x = "Closing Price", y = "Frequency") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold", size = 10))

# Density plot - Top 5 companies only
ggplot(top5_data, aes(x = Close, fill = Company)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Closing Prices - Top 5 Companies",
       subtitle = "Top 5 companies with highest average closing prices",
       x = "Closing Price", y = "Density") +
  theme_minimal() +
  theme(legend.position = "right")

# Box plot - Top 5 companies only
ggplot(top5_data, aes(x = reorder(Company, Close, FUN = median), 
                      y = Close, fill = Company)) +
  geom_boxplot() +
  labs(title = "Box Plot of Closing Prices - Top 5 Companies",
       subtitle = "Ordered by median closing price",
       x = "Company", y = "Closing Price") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none")

# Alternative: You can also select top 5 by MAXIMUM closing price instead of average
# Uncomment below if you prefer maximum price
# top5_by_max_price <- all_stocks %>%
#   group_by(Company) %>%
#   summarise(Max_Close = max(Close, na.rm = TRUE)) %>%
#   arrange(desc(Max_Close)) %>%
#   head(5) %>%
#   pull(Company)

# Violin plot - Top 5 companies (shows distribution better than box plot)
ggplot(top5_data, aes(x = reorder(Company, Close, FUN = median), 
                      y = Close, fill = Company)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, alpha = 0.5, outlier.shape = NA) +
  labs(title = "Distribution of Closing Prices - Top 5 Companies",
       subtitle = "Violin plot shows full distribution shape with box plot overlay",
       x = "Company", y = "Closing Price") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none")

```



## Volume distribution


```{r volume-distribution, fig.width=12, fig.height=6}
# 1. Filter out the VN30 index and find the REAL top 5 companies by volume
top5_vol_companies <- all_stocks %>%
  filter(Company != "VN30") %>% # Exclude the market index
  group_by(Company) %>%
  summarise(Avg_Volume = mean(Volume, na.rm = TRUE)) %>%
  arrange(desc(Avg_Volume)) %>%
  slice_head(n = 5) %>%
  pull(Company)

# Print the updated top 5 to the console
cat("Updated Top 5 Companies by Volume (Excluding Index):\n")
print(top5_vol_companies)

# 2. Filter for top 5 AND restrict volume to zoom in on the bulk of the data
top5_vol_data <- all_stocks %>%
  filter(Company %in% top5_vol_companies) %>%
  filter(Volume <= 40000000) # Truncate the extreme right-tail outliers

# 3. Plot the zoomed-in Volume Distribution
ggplot(top5_vol_data, aes(x = Volume, fill = Company)) +
  geom_histogram(bins = 50, color = "black", alpha = 0.7) +
  facet_wrap(~Company, scales = "free_y", ncol = 3) + 
  scale_x_continuous(
    labels = scales::comma,
    breaks = seq(0, 40000000, by = 10000000) # Create clean breaks every 10M
  ) + 
  labs(
    title = "Trading Volume Distribution: Top 5 VN30 Companies",
    subtitle = "Excludes VN30 Index | Zoomed to trading days with <= 40M shares",
    x = "Trading Volume (Shares)",
    y = "Frequency (Days)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1) # Tilts text to prevent overlap
  )
```


## summary statistic
```{r}
# Comprehensive summary by company
summary_stats <- all_stocks %>%
  group_by(Company) %>%
  summarise(
    N_Days = n(),
    Min_Price = min(Close, na.rm = TRUE),
    Max_Price = max(Close, na.rm = TRUE),
    Mean_Price = mean(Close, na.rm = TRUE),
    Median_Price = median(Close, na.rm = TRUE),
    SD_Price = sd(Close, na.rm = TRUE),
    CV_Price = sd(Close, na.rm = TRUE) / mean(Close, na.rm = TRUE),
    Mean_Volume = mean(Volume, na.rm = TRUE),
    Mean_Return = mean(Daily_Return, na.rm = TRUE),
    SD_Return = sd(Daily_Return, na.rm = TRUE),
    Min_Return = min(Daily_Return, na.rm = TRUE),
    Max_Return = max(Daily_Return, na.rm = TRUE)
  )

print(summary_stats)

# Export summary
write.csv(summary_stats, "summary_statistics.csv", row.names = FALSE)
```
That's stats information, but what's the meaning of those number?


this is too colorful!! Instead, showing highest, lowest, the most fluctuated, most increased/decreased. Or what happended in 2018-2019




# TSA 
top 5 only

## Calculate moving averages for stock_sample first
```{r}

stock_sample <- stock_sample %>%
  arrange(TradingDate) %>%
  mutate(
    MA_7 = rollmean(Close, k = 7, fill = NA, align = "right"),
    MA_30 = rollmean(Close, k = 30, fill = NA, align = "right"),
    MA_90 = rollmean(Close, k = 90, fill = NA, align = "right")
  )

```


## stock price trends
```{r tsa-price-trends, fig.width=12, fig.height=8}
library(ggplot2)
library(dplyr)
library(scales)

# 1. Ensure we have the top 5 companies (excluding VN30 index)
top5_companies <- all_stocks %>%
  filter(Company != "VN30") %>%
  group_by(Company) %>%
  summarise(Avg_Volume = mean(Volume, na.rm = TRUE)) %>%
  arrange(desc(Avg_Volume)) %>%
  slice_head(n = 5) %>%
  pull(Company)

# Filter the dataset for these top 5 companies
tsa_data <- all_stocks %>%
  filter(Company %in% top5_companies) %>%
  arrange(Company, TradingDate) # Ensure data is sorted by time

# ---------------------------------------------------------
# Plot 1: Overlaid Line Chart (All on one axis)
# ---------------------------------------------------------
p1 <- ggplot(tsa_data, aes(x = TradingDate, y = Close, color = Company)) +
  geom_line(alpha = 0.8, linewidth = 0.8) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Price Trends Over Time (Overlaid)",
    subtitle = "Closing Prices of the Top 5 Most Liquid VN30 Companies",
    x = "Trading Date",
    y = "Closing Price (VND)",
    color = "Ticker"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )

print(p1)

# ---------------------------------------------------------
# Plot 2: Faceted Line Chart (Independent Y-axes)
# ---------------------------------------------------------
p2 <- ggplot(tsa_data, aes(x = TradingDate, y = Close, color = Company)) +
  geom_line(alpha = 0.9, linewidth = 0.7) +
  facet_wrap(~Company, scales = "free_y", ncol = 2) + # free_y allows independent axes
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Price Trends Over Time (Faceted)",
    subtitle = "Independent price scales reveal individual volatility and trends better",
    x = "Trading Date",
    y = "Closing Price (VND)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none", # Hide legend since facets are labeled
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )

print(p2)
```

#   Multivariate Analysis

## Correlation Analysis
```{r}
# Select one stock for correlation analysis
stock_numeric <- stock_sample %>%
  select(Open, High, Low, Close, Volume) %>%
  na.omit()

# Correlation matrix
cor_matrix <- cor(stock_numeric)
print(cor_matrix)

# Correlation plot
corrplot(cor_matrix, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix of Stock Variables")

# Pairwise scatter plots
pairs(stock_numeric[, 1:5], pch = 19, col = rgb(0, 0, 1, 0.3),
      main = "Pairwise Relationships")
```
## Price Correlation Between Stocks

```{r}

# Create wide format for closing prices
price_wide <- all_stocks %>%
select(TradingDate, Company, Close) %>%
pivot_wider(names_from = Company, values_from = Close)
# Calculate correlation
price_cor <- cor(price_wide[, -1], use = "pairwise.complete.obs")
# Correlation heatmap
corrplot(price_cor, method = "color", type = "upper",
tl.col = "black", tl.srt = 45, tl.cex = 0.7,
title = "Correlation Between Stock Prices")
# Find highly correlated pairs
highly_correlated <- which(abs(price_cor) > 0.8 & price_cor != 1, arr.ind = TRUE)

```
