---
title: "EDA<br>Stock Prices & Volume VN30 Index Vietnam"
subtitle: "I create extra version to display all companies"

author: "**Pham_Dinh_Bao_Khoi** (Étudiant)<br>khoipdb.23bi14230@usth.edu.vn"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
    highlight: espresso
    code_download: TRUE
    code_folding: show
    number_sections: TRUE
    dev: "svg"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.dim = c(8, 6), fig.align = "center", out.width = "100%")
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(c('r', 'python', 'json', 'linux'), position = c('top', 'right'),
               tooltip_message = 'Click to copy', tooltip_success = 'Copied')
```

------------------------------------------------------------------------

# libraries

```{r}
# install.packages(c("tidyverse", "lubridate", "zoo", "quantmod","PerformanceAnalytics", "corrplot", "plotly","gridExtra", "scales", "TTR", "tseries"))



library(tidyverse)      # Data manipulation and visualization
library(lubridate)      # Date handling
library(zoo)            # TradingDate series
library(quantmod)       # Financial analysis
library(PerformanceAnalytics)  # Financial performance metrics
library(corrplot)       # Correlation plots
library(plotly)         # Interactive plots
library(gridExtra)      # Multiple plots
library(scales)         # Scale functions for visualization
library(TTR)            # Technical trading rules
library(tseries)  

```

```{r}
# list.files("ASTD/stock_price_VN30/") <--   CHANGE PATH TO FILE

library(readr)

list_of_files <- list.files(path = "/home/palm/ASTD/stock_price_VN30/",
                             recursive = TRUE,
                             pattern = "//.csv$",
                             full.names = TRUE)
dta <- readr::read_csv(list_of_files)

```

```{r}
# Set working directory
# Set working directory
setwd("/home/palm/ASTD/stock_price_VN30")

# Get list of all CSV files
csv_files <- list.files(pattern = "*.csv")

# Load all files into a list
stock_list <- lapply(csv_files, function(file) {
  df <- read.csv(file, stringsAsFactors = FALSE)
  df$Company <- tools::file_path_sans_ext(file)  # Add company name
  return(df)
})

# Combine all stocks into one dataframe
all_stocks <- bind_rows(stock_list)

# View the structure
str(all_stocks)
head(all_stocks)

```
```{r}
# Example: Load one stock for detailed analysis
stock_sample <- read.csv("/home/palm/ASTD/stock_price_VN30/MSN.csv")

head(stock_sample)


```
## basic info

```{r}
# Dimensions
dim(all_stocks)
dim(stock_sample)

# Column names
colnames(all_stocks)

# Data types
str(all_stocks)

# First and last rows
head(all_stocks, 10)
tail(all_stocks, 10)

# Summary statistics
summary(all_stocks)

# Unique companies
unique(all_stocks$Company)
length(unique(all_stocks$Company))


```
## check Missing values
```{r}
# Missing values per column
colSums(is.na(all_stocks))

# Percentage of missing values
missing_pct <- colMeans(is.na(all_stocks)) * 100
print(missing_pct) # it show blank figure

# Visualize missing values
library(naniar)
vis_miss(all_stocks)

# Missing values by company
all_stocks %>%
  group_by(Company) %>%
  summarise(
    total_rows = n(),
    missing_close = sum(is.na(Close)),
    missing_volume = sum(is.na(Volume))
  )

```
## Check for Duplicates
however, we can easily see that, there are no uplicated values
this dataset so CLEAN !!

```{r}
# Check for duplicate rows
duplicates <- all_stocks %>%
  group_by(Company, TradingDate) %>%
  filter(n() > 1)

nrow(duplicates)

# Remove duplicates if any
all_stocks <- all_stocks %>%
  distinct(Company, TradingDate, .keep_all = TRUE)
```


## Create Additional Features

explain how this section works, why it's necessary

```{r}
# Daily returns (percentage change)
all_stocks <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  mutate(
    Daily_Return = (Close - lag(Close)) / lag(Close) * 100,
    Log_Return = log(Close / lag(Close)),
    Price_Change = Close - lag(Close),
    Volume_Change = Volume - lag(Volume),
    High_Low_Range = High - Low,
    High_Low_Pct = (High - Low) / Low * 100
  ) %>%
  ungroup()

# Price range features - update 21/02/2026
all_stocks <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  mutate(
    Price_Change = Close - lag(Close),
    Volume_Change = Volume - lag(Volume),
    High_Low_Range = High - Low,
    High_Low_Pct = (High - Low) / Low * 100
  ) %>%
  ungroup()

# Moving averages
all_stocks <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  mutate(
    MA_7 = rollmean(Close, k = 7, fill = NA, align = "right"),
    MA_30 = rollmean(Close, k = 30, fill = NA, align = "right"),
    MA_90 = rollmean(Close, k = 90, fill = NA, align = "right")
  ) %>%
  ungroup()

# Volatility (rolling standard deviation)
all_stocks <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  mutate(
    Volatility_30 = rollapply(Daily_Return, width = 30, 
                               FUN = sd, fill = NA, align = "right")
  ) %>%
  ungroup()

# Time-based features
all_stocks <- all_stocks %>%
  mutate(
    Year = year(TradingDate),
    Month = month(TradingDate, label = TRUE),
    Quarter = quarter(TradingDate),
    DayOfWeek = wday(TradingDate, label = TRUE),
    WeekOfYear = week(TradingDate)
  )

```


# Univariate Analysis (cont)
## Distribution of Closing Prices
Closing price in what time? Is this an avg of all time?

From this box plot, what is the insight? what information we can understand from these figures?

The Density plot is too wide/ small, how to make it more clearly?

Showing Distribution of 1 company for instance.

```{r}
# Histogram
ggplot(all_stocks, aes(x = Close)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  facet_wrap(~Company, scales = "free") +
  labs(title = "Distribution of Closing Prices by Company",
       x = "Closing Price", y = "Frequency") +
  theme_minimal()

# Density plot
ggplot(all_stocks, aes(x = Close, fill = Company)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Closing Prices",
       x = "Closing Price", y = "Density") +
  theme_minimal()

# Box plot
ggplot(all_stocks, aes(x = Company, y = Close, fill = Company)) +
  geom_boxplot() +
  labs(title = "Box Plot of Closing Prices by Company",
       x = "Company", y = "Closing Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")



```


#   Multivariate Analysis (cont)


## Price vs Volume Relationship

**CONCLUSION FOR THIS SECTION**
It is hard to analyze this figure because the differences in 'type' of company. Some of it is Banking, Fintech, real estate investment, F&B , etc...
Therefore, it is meaningless to show the correlation 



```{r}
# Scatter plot with regression
ggplot(all_stocks, aes(x = Volume, y = Close)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  facet_wrap(~Company, scales = "free") +
  labs(title = "Price vs Volume Relationship",
       x = "Trading Volume", y = "Closing Price") +
  theme_minimal()

# Calculate correlation by company
price_volume_cor <- all_stocks %>%
  group_by(Company) %>%
  summarise(Correlation = cor(Close, Volume, use = "complete.obs")) %>%
  arrange(desc(abs(Correlation)))

print(price_volume_cor)
```


# TSA 

This is a mess--> meaningless. 
Show only top 5 highest/lowest

## Prices trends over time

```{r}
# All stocks price movement
ggplot(all_stocks, aes(x = TradingDate, y = Close, color = Company)) +
  geom_line(alpha = 0.7) +
  labs(title = "Stock Price Trends - All Companies",
       x = "TradingDate", y = "Closing Price") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Individual stock with moving averages
stock_sample %>%
  ggplot(aes(x = TradingDate)) +
  geom_line(aes(y = Close, color = "Actual Price")) +
  geom_line(aes(y = MA_7, color = "MA 7")) +
  geom_line(aes(y = MA_30, color = "MA 30")) +
  geom_line(aes(y = MA_90, color = "MA 90")) +
  labs(title = "Stock Price with Moving Averages",
       x = "TradingDate", y = "Price", color = "Legend") +
  theme_minimal()

# Faceted time series
ggplot(all_stocks, aes(x = TradingDate, y = Close)) +
  geom_line(color = "steelblue") +
  facet_wrap(~Company, scales = "free_y", ncol = 5) +
  labs(title = "Individual Stock Price Trends",
       x = "TradingDate", y = "Closing Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Volume trends

```{r}
# Volume over time
ggplot(all_stocks, aes(x = TradingDate, y = Volume)) +
  geom_line(color = "coral") +
  facet_wrap(~Company, scales = "free_y", ncol = 5) +
  labs(title = "Trading Volume Trends by Company",
       x = "Trading Date", y = "Volume") +
  theme_minimal()

# Price vs Volume
ggplot(all_stocks, aes(x = TradingDate)) +
  geom_line(aes(y = Close, color = "Price")) +
  geom_bar(aes(y = Volume / max(Volume, na.rm = TRUE) * max(Close, na.rm = TRUE)), 
           stat = "identity", alpha = 0.3, fill = "gray") +
  facet_wrap(~Company, scales = "free_y", ncol = 3) +
  labs(title = "Price and Volume Trends",
       x = "Trading Date", y = "Value") +
  theme_minimal()
```

## Seasoning 

```{r}
# Monthly average prices
monthly_avg <- all_stocks %>%
  group_by(Company, Year, Month) %>%
  summarise(
    Avg_Price = mean(Close, na.rm = TRUE),
    Avg_Volume = mean(Volume, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(monthly_avg, aes(x = Month, y = Avg_Price, fill = Company)) +
  geom_col(position = "dodge") +
  labs(title = "Average Prices by Month",
       x = "Month", y = "Average Closing Price (VND)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Day of week effect on prices
dow_prices <- all_stocks %>%
  group_by(DayOfWeek) %>%
  summarise(
    Avg_Price = mean(Close, na.rm = TRUE),
    Median_Price = median(Close, na.rm = TRUE)
  )

ggplot(dow_prices, aes(x = DayOfWeek, y = Avg_Price)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average Prices by Day of Week",
       x = "Day of Week", y = "Average Price (VND)") +
  theme_minimal()

# Yearly performance
yearly_perf <- all_stocks %>%
  group_by(Company, Year) %>%
  summarise(
    Yearly_Change = (last(Close) - first(Close)) / first(Close) * 100,
    .groups = "drop"
  )

ggplot(yearly_perf, aes(x = Year, y = Yearly_Change, color = Company, group = Company)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Yearly Price Change by Company",
       x = "Year", y = "Yearly Change (%)") +
  theme_minimal()
```



#  Comparative Analysis Across Stocks

## Performance Comparison

```{r}
# Normalize prices to compare performance
normalized_prices <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  mutate(Normalized_Price = Close / first(Close) * 100) %>%
  ungroup()

ggplot(normalized_prices, aes(x = TradingDate, y = Normalized_Price, color = Company)) +
  geom_line(alpha = 0.7) +
  labs(title = "Normalized Stock Performance (Base = 100)",
       x = "Trading Date", y = "Normalized Price") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Price Range Analysis

```{r}
# Calculate price range metrics by company
price_range_analysis <- all_stocks %>%
  group_by(Company) %>%
  summarise(
    Avg_High_Low_Range = mean(High_Low_Range, na.rm = TRUE),
    Avg_Price = mean(Close, na.rm = TRUE),
    Range_Pct = Avg_High_Low_Range / Avg_Price * 100
  )

# Price range scatter plot
ggplot(price_range_analysis, aes(x = Avg_Price, y = Avg_High_Low_Range, label = Company)) +
  geom_point(size = 4, color = "steelblue") +
  geom_text(hjust = -0.1, vjust = 0.5, size = 3) +
  labs(title = "Average Price vs Daily Price Range",
       x = "Average Price (VND)", y = "Average Daily Range (VND)") +
  theme_minimal()
```

## Best and Worst Performers

```{r}
# Total price change calculation
total_performance <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  summarise(
    Start_Price = first(Close),
    End_Price = last(Close),
    Total_Change_Pct = (End_Price - Start_Price) / Start_Price * 100,
    Trading_Days = n()
  ) %>%
  arrange(desc(Total_Change_Pct))

# Bar plot
ggplot(total_performance, aes(x = reorder(Company, Total_Change_Pct), 
                          y = Total_Change_Pct, fill = Total_Change_Pct > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("red", "green"), guide = FALSE) +
  labs(title = "Total Price Change by Company",
       x = "Company", y = "Total Change (%)") +
  theme_minimal()

# Top 5 and Bottom 5
cat("Top 5 Performers:\n")
print(head(total_performance, 5))
cat("\nBottom 5 Performers:\n")
print(tail(total_performance, 5))
```




