---
title: "EDA<br>Stock Prices & Volume VN30 Index Vietnam"

author: "**Pham_Dinh_Bao_Khoi** (Étudiant)<br>khoipdb.23bi14230@usth.edu.vn"
date: "`r Sys.Date()`"
output: 
  rmdformats::readthedown:
    highlight: espresso
    code_download: TRUE
    code_folding: show
    number_sections: TRUE
    dev: "svg"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.dim = c(8, 6), fig.align = "center", out.width = "100%")
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(c('r', 'python', 'json', 'linux'), position = c('top', 'right'),
               tooltip_message = 'Click to copy', tooltip_success = 'Copied')
```

------------------------------------------------------------------------
# libraries

```{r}
# install.packages(c("tidyverse", "lubridate", "zoo", "quantmod","PerformanceAnalytics", "corrplot", "plotly","gridExtra", "scales", "TTR", "tseries"))



library(tidyverse)      # Data manipulation and visualization
library(lubridate)      # Date handling
library(zoo)            # TradingDate series
library(quantmod)       # Financial analysis
library(PerformanceAnalytics)  # Financial performance metrics
library(corrplot)       # Correlation plots
library(plotly)         # Interactive plots
library(gridExtra)      # Multiple plots
library(scales)         # Scale functions for visualization
library(TTR)            # Technical trading rules
library(tseries)  

```

```{r}
# list.files("ASTD/stock_price_VN30/") <--   CHANGE PATH TO FILE

library(readr)

list_of_files <- list.files(path = "/home/palm/ASTD/stock_price_VN30/",
                             recursive = TRUE,
                             pattern = "//.csv$",
                             full.names = TRUE)
dta <- readr::read_csv(list_of_files)

```

```{r}
# Set working directory
# Set working directory
setwd("/home/palm/ASTD/stock_price_VN30")

# Get list of all CSV files
csv_files <- list.files(pattern = "*.csv")

# Load all files into a list
stock_list <- lapply(csv_files, function(file) {
  df <- read.csv(file, stringsAsFactors = FALSE)
  df$Company <- tools::file_path_sans_ext(file)  # Add company name
  return(df)
})

# Combine all stocks into one dataframe
all_stocks <- bind_rows(stock_list)

# View the structure
str(all_stocks)
head(all_stocks)

```
```{r}
# Example: Load one stock for detailed analysis
stock_sample <- read.csv("/home/palm/ASTD/stock_price_VN30/MSN.csv")

head(stock_sample)


```


```{r}
# Dimensions
dim(all_stocks)
dim(stock_sample)

# Column names
colnames(all_stocks)

# Data types
str(all_stocks)

# First and last rows
head(all_stocks, 10)
tail(all_stocks, 10)

# Summary statistics
summary(all_stocks)

# Unique companies
unique(all_stocks$Company)
length(unique(all_stocks$Company))


```

```{r}
# Missing values per column
colSums(is.na(all_stocks))

# Percentage of missing values
missing_pct <- colMeans(is.na(all_stocks)) * 100
print(missing_pct) # it show blank figure

# Visualize missing values
library(naniar)
vis_miss(all_stocks)

# Missing values by company
all_stocks %>%
  group_by(Company) %>%
  summarise(
    total_rows = n(),
    missing_close = sum(is.na(Close)),
    missing_volume = sum(is.na(Volume))
  )

```
## Check for Duplicates
however, we can easily see that, there are no uplicated values
this dataset so CLEAN !!

```{r}
# Check for duplicate rows
duplicates <- all_stocks %>%
  group_by(Company, TradingDate) %>%
  filter(n() > 1)

nrow(duplicates)

# Remove duplicates if any
all_stocks <- all_stocks %>%
  distinct(Company, TradingDate, .keep_all = TRUE)
```

# Data Cleaning and Preprocessing

## Cleaning missing values
```{r}
all_stocks <- all_stocks %>%
  select(where(~ !all(is.na(.))))

```



## DAte conversion
```{r}
# Convert Date column to Date type
all_stocks$TradingDate <- dmy(all_stocks$TradingDate)

# Sort by company and date
all_stocks <- all_stocks %>%
  arrange(Company, TradingDate)

# Check date range
range(all_stocks$TradingDate, na.rm = TRUE)
```

## Create Additional Features

explain how this section works, why it's necessary

```{r}
# Daily returns (percentage change)
all_stocks <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  mutate(
    Daily_Return = (Close - lag(Close)) / lag(Close) * 100,
    Log_Return = log(Close / lag(Close)),
    Price_Change = Close - lag(Close),
    Volume_Change = Volume - lag(Volume),
    High_Low_Range = High - Low,
    High_Low_Pct = (High - Low) / Low * 100
  ) %>%
  ungroup()

# Moving averages
all_stocks <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  mutate(
    MA_7 = rollmean(Close, k = 7, fill = NA, align = "right"),
    MA_30 = rollmean(Close, k = 30, fill = NA, align = "right"),
    MA_90 = rollmean(Close, k = 90, fill = NA, align = "right")
  ) %>%
  ungroup()

# Volatility (rolling standard deviation)
all_stocks <- all_stocks %>%
  group_by(Company) %>%
  arrange(TradingDate) %>%
  mutate(
    Volatility_30 = rollapply(Daily_Return, width = 30, 
                               FUN = sd, fill = NA, align = "right")
  ) %>%
  ungroup()

# Time-based features
all_stocks <- all_stocks %>%
  mutate(
    Year = year(TradingDate),
    Month = month(TradingDate, label = TRUE),
    Quarter = quarter(TradingDate),
    DayOfWeek = wday(TradingDate, label = TRUE),
    WeekOfYear = week(TradingDate)
  )

```

# Univariate Analysis

## Distribution of Closing Prices
Closing price in what time? Is this an avg of all time?

From this box plot, what is the insight? what information we can understand from these figures?

The Density plot is too wide/ small, how to make it more clearly?

Showing Distribution of 1 company for instance.

```{r}
# Histogram
ggplot(all_stocks, aes(x = Close)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black") +
  facet_wrap(~Company, scales = "free") +
  labs(title = "Distribution of Closing Prices by Company",
       x = "Closing Price", y = "Frequency") +
  theme_minimal()

# Density plot
ggplot(all_stocks, aes(x = Close, fill = Company)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density Plot of Closing Prices",
       x = "Closing Price", y = "Density") +
  theme_minimal()

# Box plot
ggplot(all_stocks, aes(x = Company, y = Close, fill = Company)) +
  geom_boxplot() +
  labs(title = "Box Plot of Closing Prices by Company",
       x = "Company", y = "Closing Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

```

## Volume distribution
```{r}

ggplot(all_stocks, aes(x = Volume)) +
  geom_histogram(bins = 50, fill = "coral", color = "black") +
  facet_wrap(~Company, scales = "free") +
  labs(title = "Distribution of Trading Volume by Company",
       x = "Volume", y = "Frequency") +
  theme_minimal()

# Average volume by company
avg_volume <- all_stocks %>%
  group_by(Company) %>%
  summarise(Avg_Volume = mean(Volume, na.rm = TRUE)) %>%
  arrange(desc(Avg_Volume))

ggplot(avg_volume, aes(x = reorder(Company, Avg_Volume), y = Avg_Volume, fill = Company)) +
  geom_col() +
  coord_flip() +
  labs(title = "Average Trading Volume by Company",
       x = "Company", y = "Average Volume") +
  theme_minimal() +
  theme(legend.position = "none")
```
## daily return stats
```{r}
# Histogram of returns
ggplot(all_stocks %>% filter(!is.na(Daily_Return)), 
       aes(x = Daily_Return)) +
  geom_histogram(bins = 100, fill = "darkgreen", color = "black") +
  labs(title = "Distribution of Daily Returns",
       x = "Daily Return (%)", y = "Frequency") +
  theme_minimal()

# QQ plot to check normality
qqnorm(all_stocks$Daily_Return[!is.na(all_stocks$Daily_Return)])
qqline(all_stocks$Daily_Return[!is.na(all_stocks$Daily_Return)], col = "red")

# Statistical tests for normality
shapiro.test(sample(all_stocks$Daily_Return[!is.na(all_stocks$Daily_Return)], 5000))
```




## summary statistic
```{r}
# Comprehensive summary by company
summary_stats <- all_stocks %>%
  group_by(Company) %>%
  summarise(
    N_Days = n(),
    Min_Price = min(Close, na.rm = TRUE),
    Max_Price = max(Close, na.rm = TRUE),
    Mean_Price = mean(Close, na.rm = TRUE),
    Median_Price = median(Close, na.rm = TRUE),
    SD_Price = sd(Close, na.rm = TRUE),
    CV_Price = sd(Close, na.rm = TRUE) / mean(Close, na.rm = TRUE),
    Mean_Volume = mean(Volume, na.rm = TRUE),
    Mean_Return = mean(Daily_Return, na.rm = TRUE),
    SD_Return = sd(Daily_Return, na.rm = TRUE),
    Min_Return = min(Daily_Return, na.rm = TRUE),
    Max_Return = max(Daily_Return, na.rm = TRUE)
  )

print(summary_stats)

# Export summary
write.csv(summary_stats, "summary_statistics.csv", row.names = FALSE)
```
That's stats information, but what's the meaning of those number?


this is too colorful!! Instead, showing highest, lowest, the most fluctuated, most increased/decreased. Or what happended in 2018-2019

```{r}
# Calculate moving averages for stock_sample first
stock_sample <- stock_sample %>%
  arrange(TradingDate) %>%
  mutate(
    MA_7 = rollmean(Close, k = 7, fill = NA, align = "right"),
    MA_30 = rollmean(Close, k = 30, fill = NA, align = "right"),
    MA_90 = rollmean(Close, k = 90, fill = NA, align = "right")
  )

```


# TSA 

## Prices trends over time

```{r}
# All stocks price movement
ggplot(all_stocks, aes(x = TradingDate, y = Close, color = Company)) +
  geom_line(alpha = 0.7) +
  labs(title = "Stock Price Trends - All Companies",
       x = "TradingDate", y = "Closing Price") +
  theme_minimal() +
  theme(legend.position = "bottom")

# Individual stock with moving averages
stock_sample %>%
  ggplot(aes(x = TradingDate)) +
  geom_line(aes(y = Close, color = "Actual Price")) +
  geom_line(aes(y = MA_7, color = "MA 7")) +
  geom_line(aes(y = MA_30, color = "MA 30")) +
  geom_line(aes(y = MA_90, color = "MA 90")) +
  labs(title = "Stock Price with Moving Averages",
       x = "TradingDate", y = "Price", color = "Legend") +
  theme_minimal()

# Faceted time series
ggplot(all_stocks, aes(x = TradingDate, y = Close)) +
  geom_line(color = "steelblue") +
  facet_wrap(~Company, scales = "free_y", ncol = 5) +
  labs(title = "Individual Stock Price Trends",
       x = "TradingDate", y = "Closing Price") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Volume trends

```{r}
# Volume over time
ggplot(all_stocks, aes(x = TradingDate, y = Volume)) +
  geom_line(color = "coral") +
  facet_wrap(~Company, scales = "free_y", ncol = 5) +
  labs(title = "Trading Volume Trends by Company",
       x = "Trading Date", y = "Volume") +
  theme_minimal()

# Price vs Volume
ggplot(all_stocks, aes(x = TradingDate)) +
  geom_line(aes(y = Close, color = "Price")) +
  geom_bar(aes(y = Volume / max(Volume, na.rm = TRUE) * max(Close, na.rm = TRUE)), 
           stat = "identity", alpha = 0.3, fill = "gray") +
  facet_wrap(~Company, scales = "free_y", ncol = 3) +
  labs(title = "Price and Volume Trends",
       x = "Trading Date", y = "Value") +
  theme_minimal()



```
## seasonality analytics

```{r}
# Monthly average returns
monthly_returns <- all_stocks %>%
  group_by(Company, Month) %>%
  summarise(Avg_Return = mean(Daily_Return, na.rm = TRUE), .groups = "drop")

ggplot(monthly_returns, aes(x = Month, y = Avg_Return, fill = Company)) +
  geom_col(position = "dodge") +
  labs(title = "Average Returns by Month",
       x = "Month", y = "Average Daily Return (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Day of week effect
dow_returns <- all_stocks %>%
  group_by(DayOfWeek) %>%
  summarise(Avg_Return = mean(Daily_Return, na.rm = TRUE),
            Median_Return = median(Daily_Return, na.rm = TRUE))

ggplot(dow_returns, aes(x = DayOfWeek, y = Avg_Return)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average Returns by Day of Week",
       x = "Day of Week", y = "Average Return (%)") +
  theme_minimal()

# Yearly performance
yearly_perf <- all_stocks %>%
  group_by(Company, Year) %>%
  summarise(
    Yearly_Return = (last(Close) - first(Close)) / first(Close) * 100,
    .groups = "drop"
  )

ggplot(yearly_perf, aes(x = Year, y = Yearly_Return, color = Company, group = Company)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(title = "Yearly Returns by Company",
       x = "Year", y = "Yearly Return (%)") +
  theme_minimal()


```
## violatillity 

```{r}

# Rolling volatility
ggplot(all_stocks %>% filter(!is.na(Volatility_30)), 
       aes(x = TradingDate, y = Volatility_30, color = Company)) +
  geom_line(alpha = 0.7) +
  labs(title = "30-Day Rolling Volatility",
       x = "Trading Date", y = "Volatility (%)") +
  theme_minimal()

# Volatility ranking
vol_ranking <- all_stocks %>%
  group_by(Company) %>%
  summarise(Avg_Volatility = mean(Volatility_30, na.rm = TRUE)) %>%
  arrange(desc(Avg_Volatility))

ggplot(vol_ranking, aes(x = reorder(Company, Avg_Volatility), 
                        y = Avg_Volatility, fill = Company)) +
  geom_col() +
  coord_flip() +
  labs(title = "Average Volatility by Company",
       x = "Company", y = "Average 30-Day Volatility (%)") +
  theme_minimal() +
  theme(legend.position = "none")
```
#   Multivariate Analysis

## Correlation Analysis
```{r}
# Select one stock for correlation analysis
stock_numeric <- stock_sample %>%
  select(Open, High, Low, Close, Volume) %>%
  na.omit()

# Correlation matrix
cor_matrix <- cor(stock_numeric)
print(cor_matrix)

# Correlation plot
corrplot(cor_matrix, method = "color", type = "upper", 
         addCoef.col = "black", tl.col = "black", tl.srt = 45,
         title = "Correlation Matrix of Stock Variables")

# Pairwise scatter plots
pairs(stock_numeric[, 1:5], pch = 19, col = rgb(0, 0, 1, 0.3),
      main = "Pairwise Relationships")
```
## Price Correlation Between Stocks

```{r}

# Create wide format for closing prices
price_wide <- all_stocks %>%
  select(TradingDate, Company, Close) %>%
  pivot_wider(names_from = Company, values_from = Close)

# Calculate correlation
price_cor <- cor(price_wide[, -1], use = "pairwise.complete.obs")

# Correlation heatmap
corrplot(price_cor, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45, tl.cex = 0.7,
         title = "Correlation Between Stock Prices")

# Find highly correlated pairs
highly_correlated <- which(abs(price_cor) > 0.8 & price_cor != 1, arr.ind = TRUE)
```

**CONCLUSION FOR THIS SECTION**
It is hard to analyze this figure because the differences in 'type' of company. Some of it is Banking, Fintech, real estate investment, F&B , etc...
Therefore, it is meaningless to show the correlation 

